---
title: "acciddasuite"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{acciddasuite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 7,
  dpi = 300
)
```

## Overview

`acciddasuite` builds infectious disease forecasts in a few steps:

1. **`get_data()`** or **`check_data()`**: fetch or validate surveillance data.
2. **`get_ncast()`** *(optional)*: correct recent weeks for reporting delays.
3. **`get_cv()`** *(optional)*: evaluate candidate models by time series cross-validation.
4. **`get_fcast()`**: ensemble the best models into a forward-looking forecast.

The package relies on the [`fable`](https://fable.tidyverts.org/) modeling framework and follows the standard forecasting workflow described by [Hyndman & Athanasopoulos (2021)](https://otexts.com/fpp3/basic-steps.html). The overall goal is to provide public health professionals with an easily-adoptable approach to generating, evaluating forecasts, and visualizing infectious disease forecasts.

To get more information about how to know whether forecasting is the best approach for your task, follow the steps in [this](forecast_planning.html) article.

## Step 1: Get data

We fetch weekly COVID-19 hospital admissions for New York from the [CDC NHSN](https://data.cdc.gov/Public-Health-Surveillance/Weekly-Hospital-Respiratory-Data-HRD-Metrics-by-Ju/mpgq-jmmr/about_data) via [`epidatr`](https://cmu-delphi.github.io/epidatr/).

Setting `revisions = TRUE` retrieves the full revision history (*i.e.* all past versions of the data), which is needed for nowcasting.

`get_data()` returns a validated `accidda_data` object:

```{r, get_data}
library(acciddasuite)
df <- get_data(pathogen = "covid", geo_value = "ny", revisions = TRUE)
df
```

You can also **bring your own data**. Just pass it through `check_data()`. See `vignette("external_data")` for formatting details.

## Step 2: Nowcasting (optional)

The most recent weeks of surveillance data are almost always too low because hospitals are still filing late reports (**right truncated**). If you feed these raw counts into a forecaster, predictions will be biased downward.

`get_ncast()` estimates what the recent counts will look like once all reports arrive. With the default `max_delay = 4`, the last 4 weeks are corrected; everything before that is left untouched.

```{r, nowcast}
ncast <- get_ncast(df)
ncast
```

```{r, plot-nowcast}
ncast$plot
```

The corrected `ncast$data` contains two extra columns: `ncast_lower` and `ncast_upper` (95% CrI) for the corrected weeks. `get_fcast()` detects these automatically and uses them to propagate nowcasting uncertainty into the final forecast.

## Step 3: Forecasting

Forecasting is split into two steps:

1. **`get_cv()` — model selection**: time series cross-validation on the full (median corrected) series, starting from `eval_start_date`. Models are ranked by WIS and interval coverage.
2. **`get_fcast()` — final forecast**: reuses the ranking to ensemble the best `top_n` models and projects `h` weeks into the future. When nowcast columns are present, the forecast is produced from three baselines (lower, median, and upper nowcast estimates) and pooled, so prediction intervals reflect both model uncertainty and nowcast uncertainty.

We set `eval_start_date` to mark the start of the evaluation window. At least 52 weeks of data must precede this date.

```{r, tscv-setup}
eval_start_date <- max(ncast$data$target_end_date) - 28
```

Default models are:

* `NAIVE` (Naïve / random walk): Carries the last observed value forward. The simplest possible baseline.

* `ETS` (Exponential Smoothing): A weighted average where recent weeks matter more than older ones. Adapts to trends and seasonal patterns.

* `THETA`: Splits the data into a long-term trend and short-term fluctuations, forecasts each separately, then combines them.

* `ARIMA`: Learns repeating patterns from past values to predict future ones. Auto-configured to find the best fit.

```{r, cross-validation}
cv <- get_cv(
  ncast,
  eval_start_date = eval_start_date,
  h = 4
)
cv

```

```{r, final-forecast}
fcast <- get_fcast(cv, top_n = 3)
fcast
```


### Adding custom models

Any model compatible with the [`fable`](https://fable.tidyverts.org/) framework can be passed to `get_cv()` via `models`. Compose with `default_models()` to keep the built-ins alongside your own:

```{r, extra-models}
library(fable)
library(fable.prophet)
library(EpiEstim)
library(projections)
my_models <- c(
  default_models(),
  list(
    CUSTOM_ARIMA = ARIMA(observation ~ pdq(1,1,0)),
    PROPHET = prophet(observation ~ season("year")),
    EPIESTIM = EPIESTIM(observation, mean_si = 3, std_si = 2, rt_window = 7)
  )
)

cv <- get_cv(
  ncast,
  eval_start_date = eval_start_date,
  h = 3,
  models = my_models
)

fcast <- get_fcast(cv, top_n = 3)

```


## Submit to RespiLens

[RespiLens](https://www.respilens.com/) is a platform for sharing respiratory disease forecasts. Use `to_respilens()` to export the forecast as JSON for upload to [MyRespiLens](https://www.respilens.com/myrespilens).

```{r, to-respilens, eval = FALSE}
to_respilens(fcast, "respilens.json")
```
