---
title: "Ebola 2026"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ebola 2026}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 7,
  dpi = 300
)
```

## Overview

In May 2026, the Democratic Republic of the Congo (DRC) reported an Ebola outbreak caused by the Bundibugyo strain, five months after the previous epidemic. By July 2026, 2,423 cases and 967 deaths had been reported. Rapid forecasting is critical for guiding public health response and resource allocation. This vignette demonstrates how to use `incast` to forecast Ebola incidence using surveillance data from the [Institut National de Recherche Biomédicale (INRB)](https://github.com/INRB-UMIE/BDBV2026-Data).

The vignette is not executed because the models are computationally expensive. Running the code locally may take several minutes.

## Data

We model the six locations with the highest cumulative confirmed cases. For each location, we create a daily time series up to the latest reporting date, filling missing days by carrying forward cumulative counts and removing downward corrections. Leading zeros before the first reported case are removed because the models use log-transformed incidence; locations with later introductions therefore have shorter time series.

```{r, data, eval = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

ebola <- read.csv(
  "https://raw.githubusercontent.com/scc-usc/ebola2026/refs/heads/main/hubverse_observed_data.csv"
) |>
  filter(
    target == "insp_sitrep__cumulative_confirmed_cases__daily",
    !is.na(location)
  ) |>
  mutate(target_end_date = as.Date(target_end_date))

top_locations <- ebola |>
  slice_max(target_end_date, by = location) |>
  slice_max(observation, n = 6) |>
  pull(location)

ebola <- ebola |>
  filter(location %in% top_locations) |>
  arrange(location, target_end_date) |>
  group_by(location) |>
  complete(
    target_end_date = seq(
      min(target_end_date),
      max(target_end_date),
      by = "day"
    )
  ) |>
  fill(observation, .direction = "down") |> # carry cumulative over gaps
  mutate(observation = cummax(coalesce(observation, 0))) |> # monotonic, per location
  filter(cumsum(observation > 0) > 0) |> # drop pre-first-case zeros; log() needs > 0
  ungroup() |>
  mutate(target = "insp_sitrep__cumulative_confirmed_cases__daily")
```


## `incast` Workflow

### Data Validation

`check_data()` validates the Ebola data and returns an `incast_data` object. The function checks for missing values, ensures that the data is in the correct format, and verifies that the necessary columns are present.

```{r, check_data, eval = FALSE}
library(incast)
data <- ebola |> check_data()
```

```{r, load-data, include = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(incast)
data <- readRDS("ebola_data.rds")
```

```{r, show-data}
data
data |> autoplot()
```

Revised history of cumulative confirmed cases is not available for this outbreak, so we will skip `get_ncast()` and proceed directly to cross-validation and forecasting.


### Cross Validation

`get_cv()` performs rolling-origin time series cross-validation and returns an `incast_cv` object containing the results for each model and location.

First we define a list of models to test. We will use the default models provided by `incast` from `fable` and add some custom models, including ARIMA, a neural network, and several foundation models.

```{r, models, eval = FALSE}
library(fable)
library(fable.prophet)

models <- c(
  default_models(),
  list(
    CUSTOM_ARIMA = ARIMA(log(observation) ~ pdq(1, 1, 0)),
    NNETAR = NNETAR(log(observation), n_networks = 10),
    PROPHET = prophet(log(observation)),
    # you will need a python environment (see reticulate)
    CHRONOS = FOUNDATION(log(observation), "chronos"),
    TIMESFM = FOUNDATION(log(observation), "timesfm"),
    SUNDIAL = FOUNDATION(log(observation), "sundial"),
    MOIRAI = FOUNDATION(log(observation), "moirai")
  )
)
```

Here we forecast 7 days ahead with 16 origins spaced 1 day apart, so the evaluation period spans 22 days.

```{r, cross-validation, eval = FALSE}
h <- 7 # forecast horizon
step <- 1 # one origin per day
N <- 16 # number of origins
t <- max(data$data$target_end_date) # last observation date
eval_start_date <- t - ((h - 1) + (N - 1) * step) * data$interval

cv <- data |>
  get_cv(
    eval_start_date = eval_start_date,
    h = h,
    step = step,
    models = models
  )
```

```{r, load-cv, include = FALSE}
cv <- readRDS("ebola_cv.rds")
```

```{r, show-cv}
cv
cv |>
  autoplot() +
  ggplot2::scale_x_continuous(transform = "log2")
```

`autoplot()` summarises cross-validation performance using relative WIS (`wis_relative_skill` in `cv$score`). Raw WIS depends on the scale of the observed data and is therefore not directly comparable across locations. Relative WIS normalises scores within each location, allowing performance to be compared across locations.

The log scale makes relative differences easier to interpret: values of 0.5 and 2 indicate half and double the reference WIS, respectively, and are equally distant from the reference value of 1 on the log scale.

You can also build your own summaries from `cv$score`. For example, raw WIS per model and location.

```{r, cv-custom-plot}
cv$score |>
  group_by(location) |>
  arrange(wis, .by_group = TRUE) |>
  mutate(model_id_ordered = paste(model_id, location, sep = "__")) |>
  ggplot(aes(y = reorder(model_id_ordered, wis), x = wis)) +
  facet_wrap(~location, scales = "free") +
  geom_col(aes(fill = model_id), show.legend = FALSE) +
  scale_y_discrete(labels = \(x) sub("__.*$", "", x)) +
  theme_classic() +
  labs(y = "Model", x = "WIS")
```


### Forecasting

By default `get_fcast()` will use the best 3 model for each location based on the cross-validation results. The function returns a `incast_fcast` object containing the forecasts for each location.
```{r, forecast, eval = FALSE}
fcast <- cv |> get_fcast()
```

```{r, load-fcast, include = FALSE}
fcast <- readRDS("ebola_fcast.rds")
```

```{r, show-fcast}
fcast
fcast$meta$selection
```

`autoplot()` visualizes the ensemble forecasts for each location, including the median, 50% and 95% prediction intervals.
```{r, fcast-plot}
fcast |> autoplot()
```

You can also build your own plot using `as_tibble()` to extract the forecast data and `ggplot2` to create a custom visualization.
```{r, fcast-custom-plot}
fcast |>
  as_tibble() |>
  ggplot() +
  geom_ribbon(
    aes(x = target_end_date, ymin = lower, ymax = upper, fill = model_id),
    alpha = 0.2,
    show.legend = FALSE
  ) +
  geom_line(
    aes(x = target_end_date, y = median, colour = model_id),
    show.legend = TRUE
  ) +
  facet_wrap(~location, scales = "free") +
  # ground truth
  geom_line(
    data = fcast$hub$oracle_output |>
      filter(target_end_date >= as.Date("2026-07-01")),
    aes(x = target_end_date, y = oracle_value),
    color = "black"
  ) +
  theme_classic() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )
```
