---
title: "Preparing External Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing External Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Bringing your own data

You can use any surveillance dataset with `acciddasuite` — just pass it through `check_data()` to validate and enter the pipeline.

## Required columns

Your data frame must have these 4 columns:

| Column | Type | Description |
|--------|------|-------------|
| `target_end_date` | Date | The date for which an observation is recorded |
| `observation` | numeric | The observed value |
| `location` | character | A single location identifier |
| `target` | character | A single target identifier (e.g., "inc hosp influenza") |

To enable **nowcasting** (correcting for reporting delays), add a 5th column:

| Column | Type | Description |
|--------|------|-------------|
| `as_of` | Date | The date the observation was reported/revised |

With `as_of`, the same `target_end_date` can appear multiple times (one row per revision). Without it, each `target_end_date` should appear once.

## Example

```{r, include=FALSE}
df <- data.frame(
  target_end_date = as.Date("2024-01-01") + seq(0, by = 7, length.out = 52),
  observation = rpois(52, lambda = 20),
  location = "NY",
  target = "inc hosp influenza"
)
```

```{r}
library(acciddasuite)
head(df)

checked <- check_data(df)
checked
```

