---
title: "Disease Forecast Planning for Public Health"
output: 
  rmarkdown::html_vignette:
    keep_md: false
always_allow_html: true
vignette: >
  %\VignetteIndexEntry{Disease Forecast Planning for Public Health}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Disease forecasting is an invaluable tool for the field public health, however, clearly defining the parameters of the problems you want to solve is often more difficult than the forecasting itself. The *acciddasuite* pakcage provides a comprehensive toolkit for generating forecasts with the expectation that users already have a concrete idea of what they would like to forecast. In this vignette, we provide a series of questions to aid forecasters in the discovery of the "why", "what", "where" and "how" for their question. This package and its documentation is still actively under development, and we welcome [contributions and feedback](https://github.com/ACCIDDA/acciddasuite/issues/new) from the community.


## Step 1: Why are we interested in forecasting?
First, there needs to be a clearly defined project to get started. Here are a set of questions to consider to begin this process:

1. What is the question that you are attempting to answer? Or what insights do you hope to gain?

    - Determine what you are trying to gain from the forecasting project. Specificity at this stage makes future stages easier!


2. Who is the audience, or, who will benefit from the insights?

    - Determine who will use the forecasts, and who the interpretation of them will benefit. 


3. How far into the future are you interested in forecasting? How far into the future do these insights need to be to be useful?

    - <7 days? 1-4 weeks? Full seasonal projections*?
        - *full seasonsal projections are known as "scenarios" and are different from forecasts. 
    - The time frame associated with your question determines whether or not forecasting is the best tool. Read more in Step 2...
    
<div class="planning-note">
STOP! Have you defined your forecasting problem using the questions above?
  
  - YES → Proceed to next step
  - NO → Continue defining the approach
</div>
    
    
## Step 2: Forecast vs. scenario?
Determine if your central question is strictly a forecast, or if it is more aligned with a scenario projection.

* Forecasts
    * Forecasts are concerned with what will happen in the future under current conditions/regardless of what interventions take place. From this, we can determine actionables or deploy resources, but our starting point is based off of "real life" and not an assumption (i.e., an unconditional projection).
    * e.g., I want advanced notice on the influenza hospitalization burden in the coming weeks. 
    
* Scenarios
  * Scenarios are concerned with what will happen if we take action *X*? Or, how will the future differe if *X* happens instead of *Y*? Often, with scenarios, you are comparing different outcomes on the basis of different assumptions (i.e., a conditional projection).  
  * e.g., I want to know what the predicted burden will be if 50% of the population is vaccinated against flu vs. only 30%.
  
If you want to produce projections of the future more than a few weeks at a time, your question is likely better suited for a scenario projection rather than a forecast (forecasts give us a look 1-4 weeks ahead of any given start date, scenarios show us entire seasons at a time). For more information on scenario projections, visit the [Scenario Modeling Hub](https://scenariomodelinghub.org/).

<div class="planning-note">
STOP! Have you confirmed that your question is best answered with a forecast?
  
  - YES → Proceed to next step
  - NO → Consider checking out scenario projections instead!
</div>


## Step 3: Define Your Data
Next, define what **pathogen**, **target** (data stream), and **spatial unit** are involved in forecasting project.

1. What **pathogen** are you interested in forecasting?

    - e.g., influenza, COVID-19, RSV, etc.

2. What data stream are you interested in forecasting (also known as your **target**)? 

    - Examples of common forecasting targets are "incidence of hospitalization" (of patients with a certain hospitalization), "percent of emergency department visits" (attributable to a certain pathogen), "deaths" (due to a certain pathogen), "hospital bed occupancy" (by patients with a certain pathogen). All of these targets are measures of disease burden that give public health professionals an idea of how much disease a population bears at a given time. The most common forecasting target for respiratory illnesses is "incidence of hospitalization", which makes it the easiest to find data on. Presently, *acciddasuite* only forecasts targets "incidence of hospitalization" and "death". 
  
3. What **spatial unit** will provide the best insight? Is there data available at that scale? 

    - e.g., national, state, county, city, health jurisdiction, hospital system, or even facility (e.g., hospital). The more granular the spatial unit, the more difficult it is to find data, so there is often a trade-off between data specificity and availability. 
    
Note that you can forecast multiple locations (e.g., multiple states or health jurisdictions) at once, but if you want forecast multiple pathogens or targets, it is best to separate those into their own distinct forecasts.

<div class="planning-note">
STOP! Have you defined your pathogen, target, and spatial unit?
  
  - YES → Proceed to next step
  - NO → Continue defining these data elements
</div>


## Step 4: Data Availability & Limitations
In our context, forecasts are mathematical predictions of a few weeks ahead given a starting point of "ground truth" data (i.e, information on what has already happened). Because of this, you need to provide forecasting models with ground truth data that matches the resolution of your forecasting question (i.e., same pathogen, same target, same locations). Data availability is often a limiting factor when considering a forecasting question. In this step, we provide a decision tree approach to determine if you can forecast with ground truth data that already exists, if you need to provide specialized data to complete your forecast, or if there is already forecasting infrastructure that answers your forecasting question.

<style>
.planning-flowchart {
  margin: 1.5rem 0;
  padding: 1rem;
  border: 1px solid #d9e3ec;
  border-radius: 1rem;
  background: linear-gradient(180deg, #fbfdff 0%, #f4f8fb 100%);
  box-shadow: 0 10px 30px rgba(22, 40, 59, 0.06);
}

.planning-lead {
  margin: 0 0 1rem 0;
  font-size: 1.05rem;
  color: #35516b;
}

.planning-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(250px, 1fr));
  gap: 1rem;
  margin: 1.25rem 0 1.75rem 0;
}

.planning-card {
  padding: 1rem 1.1rem;
  border: 1px solid #d9e3ec;
  border-radius: 1rem;
  background: #ffffff;
  box-shadow: 0 8px 24px rgba(22, 40, 59, 0.05);
}

.planning-card h4 {
  margin-top: 0;
  margin-bottom: 0.75rem;
  color: #17324d;
}

.planning-card p:last-child,
.planning-card ul:last-child {
  margin-bottom: 0;
}

.planning-card ul {
  padding-left: 1.1rem;
}

.planning-note {
  padding: 0.9rem 1rem;
  border-left: 4px solid #0d6efd;
  border-radius: 0.75rem;
  background: #eef6ff;
  color: #28435c;
}

.planning-flowchart .nodeLabel,
.planning-flowchart .nodeLabel p,
.planning-flowchart .label text,
.planning-flowchart foreignObject div {
  text-align: center !important;
}
</style>

### Data availability decision tree

<div class="planning-flowchart">

```{r, echo=FALSE}
#| fig.width: 18
#| fig.height: 8
#| out.width: "100%"
#| fig.align: "center"

DiagrammeR::mermaid("
graph TB

    A[You are here!] --> B{Are you forecasting<br/>respiratory illness at a<br/>national, state, or<br/>county/metro area resolution?}

    B -->|Yes| D[Review the main<br/>forecasting hubs]
    B -->|No| E{Do you have a ground truth<br/>data stream that matches<br/>your forecasting question?}

    D --> F{Does an existing forecast hub<br/>meet your forecasting needs?}
    F -->|Yes| I[Use those forecasts!<br/>You do not need to<br/>create your own.]
    F -->|No| K

    E -->|Yes| G{Is there enough data to<br/>generate forecasts?}
    E -->|No| H{Does the data exist,<br/>but you just don't<br/>have access to it?}

    G -->|Yes| K[Proceed to Figure 2:<br/>Data collection workflow]
    H -->|Yes| K
    G -->|No| L[Return to Step 1<br/>and re-evaluate your<br/>forecasting question]
    H -->|No| L

    classDef question fill:#fff7db,stroke:#c99600,color:#533f03,stroke-width:2px;
    classDef action fill:#eef6ff,stroke:#6ea8fe,color:#17324d,stroke-width:2px;
    classDef outcome fill:#edf8ee,stroke:#3e8f55,color:#1e4d2b,stroke-width:2px;
    classDef caution fill:#fff0f0,stroke:#c45b5b,color:#6b1f1f,stroke-width:2px;
    class A,B,E,F,G,H question;
    class D,I action;
    class K outcome;
    class L caution;
")
```

 </div>

<div class="planning-grid">
  <div class="planning-card">
    <h4>Path 1: Reuse an existing forecast</h4>
    <p>Choose this route when a public forecasting hub already covers your pathogen and geography.</p>
    <ul>
      <li>Review existing state-level forecast hubs.</li>
      <li>Check whether the forecast answers your actual planning question.</li>
      <li>If it does, you can move directly into interpretation and use.</li>
    </ul>
  </div>
  <div class="planning-card">
    <h4>Path 2: Build a local workflow</h4>
    <p>Choose this route when you need a custom target, geography, or reporting cadence.</p>
    <ul>
      <li>Confirm that data exist for the target, spatial scale, and time resolution you need.</li>
      <li>Check whether historical data are available or whether access is the main barrier.</li>
      <li>If data are not available, pause and re-evaluate the project design before modeling.</li>
    </ul>
  </div>
</div>

**Figure 2. Data collection and reporting workflow** describes the process and necessary steps to collect and organize the required data to conduct a forecast.  

<div class="planning-lead">
After you decide to build a local forecast, the next question is whether your data stream is stable enough to support a repeatable workflow.
</div>

<div class="planning-flowchart">

```{r, echo=FALSE}
#| fig.width: 16
#| fig.height: 8
#| out.width: "100%"
#| fig.align: "center"

DiagrammeR::mermaid("
graph TB

    START[From Figure 1:<br/>Begin local data workflow] --> STEP1[Collect and store<br/>available data]

    STEP1 --> CONSIST{Are data reported consistently,<br/>such as on the same day each week?}

    CONSIST -->|No| FIX[Identify reporting patterns<br/>and inconsistencies]
    FIX --> CONSIST

    CONSIST -->|Yes| TIMELINE[Define a reporting timeline<br/>that matches the data frequency]
    TIMELINE --> PIPE[Download the full dataset regularly,<br/>not only the current period]

    PIPE --> DELAY{Are there reporting delays<br/>or completeness issues?}
    DELAY -->|Yes| NOWCAST[Plan for nowcasting]
    DELAY -->|No| READY[Proceed to data formatting<br/>for forecasting]

    classDef question fill:#fff7db,stroke:#c99600,color:#533f03,stroke-width:2px;
    classDef action fill:#eef6ff,stroke:#6ea8fe,color:#17324d,stroke-width:2px;
    classDef outcome fill:#edf8ee,stroke:#3e8f55,color:#1e4d2b,stroke-width:2px;
    class START,STEP1,FIX,TIMELINE,PIPE action;
    class CONSIST,DELAY question;
    class NOWCAST,READY outcome;
")
```

 </div>

<div class="planning-grid">
  <div class="planning-card">
    <h4>What to document</h4>
    <ul>
      <li>How often the source updates.</li>
      <li>Whether old records are revised after first publication.</li>
      <li>Which files or endpoints provide full historical backfill.</li>
    </ul>
  </div>
  <div class="planning-card">
    <h4>Common warning signs</h4>
    <ul>
      <li>Reports arrive on different days from week to week.</li>
      <li>Only the most recent period is available for download.</li>
      <li>Recent observations are incomplete and need later revision.</li>
    </ul>
  </div>
</div>

## National and State Level Forecasting Hubs:
FluSight [guide](https://happygitwithr.com/https-pat) 

MetroCast [website](information.https://docs.github.com/en/authentication/keeping-your-account-and-data-secure/managing-your-personal-access-tokens)

A full list of real-time collaborative public health hubs can be found [here](https://hubverse.io/community/hubs.html#real-time-collaborative-public-health-hubs).


## Next steps:

* For NHSN data, return to Get Started and use the get_data function to run forecasts 
* To forecast local surveillance data, follow [these](external_data.html) steps for formatting.


For more information, see the documentation and help pages.
