---
title: "Disease Forecast Planning for Public Health"
output: 
  rmarkdown::html_vignette:
    keep_md: false
always_allow_html: true
vignette: >
  %\VignetteIndexEntry{Disease Forecast Planning for Public Health}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Disease forecasting is an important tool for public health, however, to maximize the utility of these tools, thoughtful planning is required to clearly define the approach to the problem needing to be addressed. While the *ACCIDDA Forecast Suite* provides a comprehensive toolkit for building infectious disease forecasting pipelines and hubs, discussions and planning prior to starting your forecasting project is important. We provide a series of questions to consider addressing below prior to using this package. This package and suite are still under active development, and we welcome contributions and feedback from the community.


## Step 1: Why are we interested in forecasting?
First, there needs to be a clearly defined project to get started. Here are a set of questions to consider to begin this process:

1. What is the question that you are attempting to answer? Or what insights do you hope to gain?

    - Determine what it is you are trying to gain from the forecasting project. Being specific here will be useful in determining your approach.


2. Who is the audience or who will benefit from the insights?

    - Documenting who will use the forecasts and how will assist in the interpretation of the forecasting output.


3. How far into the future are you interested in forecasting? 
      How far into the future do these insights need to be to be useful?
          What aspects of a forecast need to be accurate?

    - Clearly defining what you think is going to be useful information will determine how to approach the problems and if forecasting is the best tool.
    
    - For example, seasonal influenza forecasting typically involves predictions 1 to 4 weeks into the future. Forecasts provide a range of possible trajectories for that time period. Predicting a specific number, such as the number of total hospitalizations during that time period would require a different approach and models.


Answer prior to proceeding:

**Have you defined your forecasting project and approach?**

    - YES → Proceed to next steps

    - NO  → Continue defining the approach


## Step 2: Define Your Data
Next, defining what pathogen, the target (time series data), the geographical area, and the time resolution in order to know what data is required for the forecasting project.

1. What *pathogen(s)* are you interested in?

While straight-forward and possibly answered in the step above, ensuring you are clear on which pathogen (e.g., influenza, COVID-19, RSV, etc) is being forecasted is important.

2. What outcomes do you measure for this pathogen?

3. Which of these outcomes is most relevant to the questions, insights, and your audience? 

4. These outcomes are also referred to as the *target*. More importantly, how timely is this target reported?

    - Is the timing of the target adequate enough for updating forecasts and evaluation? 

Determining what is specifically being forecasted will important in the data collection steps. 

Some examples of forecasting targets:

    - Reported cases of influenza
    - COVID-19 Hospitalizations
    - Respiratory disease deaths
    - Emergency Department (ED) visits related to respiratory disease-like illnesses
  
5. What *spatial unit* will provide the best insight? Are these data available at that scale? 

This may include, but are not limited to:

    - State, county, city, health jurisdiction, hospital system, or even facility (e.g., hospital)
  
6. What *time resolution* is adaquate and available to provide the required information for the audience? 

Planning the time steps is important for determining if your data is consistently and readily available for that resolution. Also, this will assist in thinking about the reporting delays or lag time for each of these time steps. 

Some typical forecasting time steps include days, weeks, or even months. 


Answer prior to proceeding:

**Have you defined your pathogen, target, spatial unit, and time resolution?**

    - YES → Proceed to next steps

    - NO  → Continue defining these data elements


## Step 3: Data Availability & Limitations
In this step, we provide a decision tree approach to assist with directing you toward a forecasting approach using available data and considering the data limitations. 

<style>
.planning-flowchart {
  margin: 1.5rem 0;
  padding: 1rem;
  border: 1px solid #d9e3ec;
  border-radius: 1rem;
  background: linear-gradient(180deg, #fbfdff 0%, #f4f8fb 100%);
  box-shadow: 0 10px 30px rgba(22, 40, 59, 0.06);
}

.planning-lead {
  margin: 0 0 1rem 0;
  font-size: 1.05rem;
  color: #35516b;
}

.planning-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(250px, 1fr));
  gap: 1rem;
  margin: 1.25rem 0 1.75rem 0;
}

.planning-card {
  padding: 1rem 1.1rem;
  border: 1px solid #d9e3ec;
  border-radius: 1rem;
  background: #ffffff;
  box-shadow: 0 8px 24px rgba(22, 40, 59, 0.05);
}

.planning-card h4 {
  margin-top: 0;
  margin-bottom: 0.75rem;
  color: #17324d;
}

.planning-card p:last-child,
.planning-card ul:last-child {
  margin-bottom: 0;
}

.planning-card ul {
  padding-left: 1.1rem;
}

.planning-note {
  padding: 0.9rem 1rem;
  border-left: 4px solid #0d6efd;
  border-radius: 0.75rem;
  background: #eef6ff;
  color: #28435c;
}
</style>

### Forecasting planning decision tree

<div class="planning-note">
Use the overview below to decide whether an existing forecast already meets your needs or whether you should prepare local data for forecasting.
</div>

**Figure 1. Initial forecast selection** describes the important initial questions needed to use existing forecasting resources or prepare to collect data for conducting your forecast. 

<div class="planning-flowchart">

```{r, echo=FALSE}
#| fig.width: 18
#| fig.height: 8
#| out.width: "100%"
#| fig.align: "center"

DiagrammeR::mermaid("
graph TB

    A[Start here:<br/>Do you need a forecast?] --> B{Use an existing<br/>state-level forecast?}

    B -->|Yes| D[Review state-level<br/>forecasting hubs]
    B -->|No| E{Are data available for your<br/>target, scale, and timing?}

    D --> F{Does an existing forecast<br/>meet your needs?}
    F -->|Yes| I[Use the existing forecast]
    F -->|No| J[Continue to local<br/>data planning]

    E -->|Yes| G{Historical data available?}
    E -->|No| H{Data exist but are<br/>hard to access?}

    G -->|Yes| K[Move to Figure 2:<br/>Data collection workflow]
    H -->|Yes| K
    G -->|No| L[Return to Step 1<br/>and re-evaluate]
    H -->|No| L

    J --> K

    classDef question fill:#fff7db,stroke:#c99600,color:#533f03,stroke-width:2px;
    classDef action fill:#eef6ff,stroke:#6ea8fe,color:#17324d,stroke-width:2px;
    classDef outcome fill:#edf8ee,stroke:#3e8f55,color:#1e4d2b,stroke-width:2px;
    classDef caution fill:#fff0f0,stroke:#c45b5b,color:#6b1f1f,stroke-width:2px;
    class A,B,E,F,G,H question;
    class D,J,K action;
    class I outcome;
    class L caution;
")
```

 </div>

<div class="planning-grid">
  <div class="planning-card">
    <h4>Path 1: Reuse an existing forecast</h4>
    <p>Choose this route when a public forecasting hub already covers your pathogen and geography.</p>
    <ul>
      <li>Review existing state-level forecast hubs.</li>
      <li>Check whether the forecast answers your actual planning question.</li>
      <li>If it does, you can move directly into interpretation and use.</li>
    </ul>
  </div>
  <div class="planning-card">
    <h4>Path 2: Build a local workflow</h4>
    <p>Choose this route when you need a custom target, geography, or reporting cadence.</p>
    <ul>
      <li>Confirm that data exist for the target, spatial scale, and time resolution you need.</li>
      <li>Check whether historical data are available or whether access is the main barrier.</li>
      <li>If data are not available, pause and re-evaluate the project design before modeling.</li>
    </ul>
  </div>
</div>

**Figure 2. Data collection and reporting workflow** describes the process and necessary steps to collect and organize the required data to conduct a forecast.  

<div class="planning-lead">
After you decide to build a local forecast, the next question is whether your data stream is stable enough to support a repeatable workflow.
</div>

<div class="planning-flowchart">

```{r, echo=FALSE}
#| fig.width: 16
#| fig.height: 8
#| out.width: "100%"
#| fig.align: "center"

DiagrammeR::mermaid("
graph TB

    START[From Figure 1:<br/>Begin local data workflow] --> STEP1[Collect and store<br/>available data]

    STEP1 --> CONSIST{Are data reported consistently,<br/>such as on the same day each week?}

    CONSIST -->|No| FIX[Identify reporting patterns<br/>and inconsistencies]
    FIX --> CONSIST

    CONSIST -->|Yes| TIMELINE[Define a reporting timeline<br/>that matches the data frequency]
    TIMELINE --> PIPE[Download the full dataset regularly,<br/>not only the current period]

    PIPE --> DELAY{Are there reporting delays<br/>or completeness issues?}
    DELAY -->|Yes| NOWCAST[Plan for nowcasting]
    DELAY -->|No| READY[Proceed to data formatting<br/>for forecasting]

    classDef question fill:#fff7db,stroke:#c99600,color:#533f03,stroke-width:2px;
    classDef action fill:#eef6ff,stroke:#6ea8fe,color:#17324d,stroke-width:2px;
    classDef outcome fill:#edf8ee,stroke:#3e8f55,color:#1e4d2b,stroke-width:2px;
    class START,STEP1,FIX,TIMELINE,PIPE action;
    class CONSIST,DELAY question;
    class NOWCAST,READY outcome;
")
```

 </div>

<div class="planning-grid">
  <div class="planning-card">
    <h4>What to document</h4>
    <ul>
      <li>How often the source updates.</li>
      <li>Whether old records are revised after first publication.</li>
      <li>Which files or endpoints provide full historical backfill.</li>
    </ul>
  </div>
  <div class="planning-card">
    <h4>Common warning signs</h4>
    <ul>
      <li>Reports arrive on different days from week to week.</li>
      <li>Only the most recent period is available for download.</li>
      <li>Recent observations are incomplete and need later revision.</li>
    </ul>
  </div>
</div>

## National and State Level Forecasting Hubs:
FluSight [guide](https://happygitwithr.com/https-pat) 

MetroCast [website](information.https://docs.github.com/en/authentication/keeping-your-account-and-data-secure/managing-your-personal-access-tokens)

A full list of real-time collaborative public health hubs can be found [here](https://hubverse.io/community/hubs.html#real-time-collaborative-public-health-hubs).


## Next steps:

* For NHSN data, return to Get Started and use the get_data function to run forecasts 
* To forecast local surveillance data, follow [these](external_data.html) steps for formatting.


For more information, see the documentation and help pages.
