---
title: "Disease Forecast Planning for Public Health"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Disease Forecast Planning for Public Health}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Disease forecasting is an important tool for public health, however, to maximize the utility of these tools, thoughtful planning is required to clearly define the approach to the problem needing to be addressed. While the *ACCIDDA Forecast Suite* provides a comprehensive toolkit for building infectious disease forecasting pipelines and hubs, discussions and planning prior to starting your forecasting project is important. We provide a series of questions to consider addressing below prior to using this package. This package and suite are still under active development, and we welcome contributions and feedback from the community.


## Step 1: Why are we interested in forecasting?
First, there needs to be a clearly defined project to get started. Here are a set of questions to consider to begin this process:

1. What is the question that you are attempting to answer? Or what insights do you hope to gain?

    - Determine what it is you are trying to gain from the forecasting project. Being specific here will be useful in determining your approach.

2. Who is the audience or who will benefit from the insights?

    - Documenting who will use the forecasts and how will assist in the interpretation of the forecasting output.
  
3. How far into the future are you interested in forecasting? 
      How far into the future do these insights need to be to be useful?
          What aspects of a forecast need to be accurate?

    - Clearly defining what you think is going to be useful information will determine how to approach the problems and if forecasting is the best tool.
    
    - For example, seasonal influenza forecasting typically involves predictions 1 to 4 weeks into the future. Forecasts provide a range of possible trajectories for that time period. Predicting a specific number, such as the number of total hospitalizations during that time period would require a different approach and models.
  
Answer prior to proceeding:

Have you defined your forecasting project and approach?

YES → Proceed to next steps

NO  → Continue defining the approach

## Step 2: Define Your Data
Next, defining what pathogen, the target (time series data), the geographical area, and the time resolution in order to know what data is required for the forecasting project.

1. What *pathogen(s)* are you interested in?

While straight-forward and possibly answered in the step above, ensuring you are clear on which pathogen (e.g., influenza, COVID-19, RSV, etc) is being forecasted is important.

2. What outcomes do you measure for this pathogen?

3. Which of these outcomes is most relevant to the questions, insights, and your audience? 

4. These outcomes are also referred to as the *target*. More importantly, how timely is this target reported?

    - Is the timing of the target adequate enough for updating forecasts and evaluation? 

Determining what is specifically being forecasted will important in the data collection steps. 

Some examples of forecasting targets:

    - Reported cases of influenza
    - COVID-19 Hospitalizations
    - Respiratory disease deaths
    - Emergency Department (ED) visits related to respiratory disease-like illnesses
  
3. What *spatial unit* will provide the best insight? Are these data available at that scale? 

This may include, but are not limited to:

    - State, county, city, health jurisdiction, hospital system, or even facility (e.g., hospital)
  
4. What *time resolution* is adaquate and available to provide the required information for the audience? 

Planning the time steps is important for determining if your data is consistently and readily available for that resolution. Also, this will assist in thinking about the reporting delays or lag time for each of these time steps. 

Some typical forecasting time steps include days, weeks, or even months. 

Answer prior to proceeding:

Have you defined your pathogen, target, spatial unit, and time resolution?

YES → Proceed to next steps

NO  → Continue defining these data elements

## Step 3: Data Availability & Limitations
In this step, you will be asked a series of questions to direct you to become familiar with the available forecasting hubs being routinely provided for national and state level targets. 


```{r eval=FALSE, include=FALSE}
#| fig-cap: acciddasuite Data Collection Flow

DiagrammeR::mermaid("
graph TB
    A[Did you choose state-level forecasts?] --> B{Yes}
    A --> C{No}
        
    B --> D[Review available forecasting hubs]
    D --> F[Did you find the information you sought?]
    
    F --> J{No}
    F --> I{Yes}
    
    C --> E[Is data available for target/spatial/time?]
    E --> G{Yes}
    E --> H{No}
    
    G --> K[Is historical data available?]
    H --> L[Data exists but not easily accessible?]
    
    K --> n1{Yes}
    K --> n2{No}
    L --> n3{Yes}
    L --> n4{No}
    
    n1 --> n18[Collect and store the available data]
    n3 --> n20[Collect any available data]
    J --> n17[Continue collecting data]
    
    n18 --> n5
    n20 --> n5
    
    n5[Are data available in a consistent manner?] --> n6{Yes}
    n5 --> n7{No}
    
    n7 --> n12[Determine reporting timeline]
    n12 --> n5
    
    n6 --> n15[Establish reporting timeline]
    n15 --> n8[Setup consistent downloads]
    
    n8 --> n9[Reporting delays or completeness issues?]
    n9 --> n10{Yes}
    n9 --> n11{No}
    
    n10 --> n13[Plan for Nowcasting]
    n11 --> n14[Proceed to data formatting]
    
    I --> n16[Congrats!]
    n2 --> n19[Reevaluate the project]
    n4 --> n21[Reevaluate the project]
")
```

```{r, echo=FALSE, include=FALSE}
#| fig.cap: "Forecasting Planning Decision Tree"

DiagrammeR::mermaid("
graph TB
    A[Did you choose state-level forecasts for a particular pathogen?] --> B{Yes} 
    
    B --> D[Review available forecasting hubs providing state-level forecasts.]
    D --> F[Did you find the state-level forecasting information you were orignially seeking?]
    
    F --> J{No} 
    F --> I{Yes}
    
    J --> n17[Continue collecting data.]
    I --> n16[Congrats!]
")
```

```{r, echo=FALSE, include=FALSE}
#| fig.cap: "Forecasting Planning Decision Tree"

DiagrammeR::mermaid("
graph TB
    A[Did you choose state-level forecasts for a particular pathogen?] --> B{Yes} 
    
    B --> D[Review available forecasting hubs providing state-level forecasts.]
    D --> F[Did you find the state-level forecasting information you were orignially seeking?]
    
    F --> J{No} 
    F --> I{Yes}
    
    J --> n17[Continue collecting data.]
    I --> n16[Congrats!]
    
    A --> C{No}
    C --> E[Is data available for the target, spactial unit, and time resolution?]
    E --> G{Yes} 
    E --> H{No}
")
```

```{r, echo=FALSE, include=FALSE}
#| fig.cap: "Forecasting Planning Decision Tree"

DiagrammeR::mermaid("
graph TB
    G{Yes} --> K[Is historical data available for the target of interest?]
    H{No} --> L[Does the data exist, but are not easily accesible?]
    
    K --> n1{Yes} 
    K --> n2{No}
    
    n1 --> n18[Collect and store the available data.]
    n2 --> n19[Go back to STEP 1. Reevaluate the project.]
    
    L --> n3{Yes}
    L --> n4{No}
    
    n3 --> n20[Collect any available data.]
    n4 --> n21[Go back to STEP 1. Reevaluate the project.]

    n18 --> n5
    n20 --> n5
    
    n5[Are these data reported or available in a consistent manner? e.g., same data is available every week on the same day] --> n6{Yes}
    n5 --> n7{No}
    
    n7 --> n12[Determine the reporting timeline and look for consistancies.]
    n12 --> n5
     
    n6 --> n15[Establish a reporting timeline consistent with the time resolution.]
    n15 --> n8
    
    n8[Setup consistent downloads of the entire dataset, not just the current week.] --> n9[Does your data have reporting delays or completeness issues?]
    n9 --> n10{Yes}
    n9 --> n11{No}
    
    n10 --> n13[Plan for Nowcasting.]
    n11 --> n14[Continue to data formatting steps for forecasting.]
    ")
```

```{r, echo=FALSE}
#| fig.cap: "Forecasting Planning Decision Tree"
#| fig.width: 14
#| fig.height: 12
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"

DiagrammeR::mermaid("
graph TB

    A[Use state-level forecasts for a pathogen?] --> B{Yes}
    A --> C{No}

    %% --- STATE FORECAST PATH ---
    B --> D[Review available state-level forecasting hubs]
    D --> F[Did you find the forecast information you need?]

    F --> I{Yes}
    F --> J{No}

    I --> END1[Done]
    J --> CONT1[Continue data collection]

    %% --- DATA AVAILABILITY PATH ---
    C --> E[Is data available at the needed scale and resolution?]
    E --> G{Yes}
    E --> H{No}

    G --> K[Is historical data available?]
    H --> L[Does the data exist but hard to access?]

    %% --- HISTORICAL DATA ---
    K --> Y1{Yes}
    K --> N1{No}

    Y1 --> STEP1[Collect and store available data]
    N1 --> RESTART1[Return to Step 1 and reassess]

    %% --- ACCESSIBILITY ---
    L --> Y2{Yes}
    L --> N2{No}

    Y2 --> STEP2[Collect any accessible data]
    N2 --> RESTART2[Return to Step 1 and reassess]

    STEP1 --> CONSIST
    STEP2 --> CONSIST

    %% --- CONSISTENCY ---
    CONSIST[Are data reported consistently?] --> CY{Yes}
    CONSIST --> CN{No}

    CN --> FIX[Identify reporting patterns and inconsistencies]
    FIX --> CONSIST

    %% --- PIPELINE SETUP ---
    CY --> TIMELINE[Define a reporting timeline]
    TIMELINE --> PIPE

    PIPE[Download the full dataset regularly--not just current data] --> DELAY
    DELAY[Are there reporting delays or missing data?] --> DY{Yes}
    DELAY --> DN{No}

    DY --> NOWCAST[Plan for nowcasting]
    DN --> READY[Proceed to data formatting for forecasting]
")
```

```{r, echo=FALSE}
#| fig.cap: "Figure 1. Forecasting planning decision tree: initial pathway selection. The data-readiness branch continues in Figure 2."
#| fig.width: 14
#| fig.height: 8
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"

DiagrammeR::mermaid("
graph TB

    A[Use state-level forecasts for a pathogen?] --> B{Yes}
    A --> C{No}

    %% --- STATE-LEVEL FORECAST PATH ---
    B --> D[Review available state-level forecasting hubs]
    D --> F[Did you find the forecast information you need?]

    F --> I{Yes}
    F --> J{No}

    I --> END1[Done]
    J --> CONT1[Continue data collection]

    %% --- DATA AVAILABILITY HANDOFF ---
    C --> E[Is data available at the needed scale and time resolution?]
    E --> G{Yes}
    E --> H{No}

    G --> K[Is historical data available?]
    H --> L[Does the data exist but is hard to access?]

    K --> Y1{Yes}
    K --> N1{No}

    L --> Y2{Yes}
    L --> N2{No}

    Y1 --> HANDOFF[Go to Figure 2: data readiness and reporting workflow]
    Y2 --> HANDOFF

    N1 --> RESTART1[Return to Step 1 and reassess]
    N2 --> RESTART2[Return to Step 1 and reassess]
")
```

## Figure 2: Data readiness and reporting workflow

```{r, echo=FALSE}
#| fig.cap: "Figure 2. Forecasting planning decision tree: data readiness and reporting workflow. This figure continues from the handoff node in Figure 1."
#| fig.width: 14
#| fig.height: 9
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"

DiagrammeR::mermaid("
graph TB

    START[From Figure 1: data readiness and reporting workflow] --> STEP1[Collect and store available data]

    STEP1 --> CONSIST[Are data reported consistently, such as on the same day each week?]

    CONSIST --> CY{Yes}
    CONSIST --> CN{No}

    CN --> FIX[Identify reporting patterns and inconsistencies]
    FIX --> CONSIST

    CY --> TIMELINE[Define a reporting timeline consistent with the data frequency]
    TIMELINE --> PIPE[Download the full dataset regularly, not only the current period]

    PIPE --> DELAY[Are there reporting delays or completeness issues?]
    DELAY --> DY{Yes}
    DELAY --> DN{No}

    DY --> NOWCAST[Plan for nowcasting]
    DN --> READY[Proceed to data formatting for forecasting]
")
```

```{r, echo=FALSE, include=FALSE}
#| fig.cap: "Forecasting Planning Decision Tree"
#| fig.width: 14
#| fig.height: 20
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"

DiagrammeR::mermaid("
graph TB

    A[Did you choose state-level forecasts for a particular pathogen?] --> B{Yes} 
    A --> C{No}

    B --> D[Review available forecasting hubs providing state-level forecasts.]
    D --> F[Did you find the state-level forecasting information you were originally seeking?]

    F --> I{Yes}
    F --> J{No} 

    I --> n16[Congrats!]
    J --> n17[Continue collecting data.]

    C --> E[Is data available for the target, spatial unit, and time resolution?]
    E --> G{Yes} 
    E --> H{No}

    G --> K[Is historical data available for the target of interest?]
    H --> L[Does the data exist, but is it not easily accessible?]

    K --> n1{Yes} 
    K --> n2{No}

    n1 --> n18[Collect and store the available data.]
    n2 --> n19[Go back to STEP 1. Reevaluate the project.]

    L --> n3{Yes}
    L --> n4{No}

    n3 --> n20[Collect any available data.]
    n4 --> n21[Go back to STEP 1. Reevaluate the project.]

    n18 --> n5
    n20 --> n5

    n5[Are these data reported or available in a consistent manner? For example, is the same data available every week on the same day?] --> n6{Yes}
    n5 --> n7{No}

    n7 --> n12[Determine the reporting timeline and look for consistencies.]
    n12 --> n5

    n6 --> n15[Establish a reporting timeline consistent with the time resolution.]
    n15 --> n8

    n8[Set up consistent downloads of the entire dataset, not just the current week.] --> n9[Does your data have reporting delays or completeness issues?]
    n9 --> n10{Yes}
    n9 --> n11{No}

    n10 --> n13[Plan for nowcasting.]
    n11 --> n14[Continue to data-formatting steps for forecasting.]
")
```

```{r, echo=FALSE, include=FALSE}
#| fig.cap: "Forecasting Planning Decision Tree"

DiagrammeR::mermaid("
graph TB
    A[Did you choose state-level forecasts for a particular pathogen?] --> B{Yes} 
    
    B --> D[Review available forecasting hubs providing state-level forecasts.]
    D --> F[Did you find the state-level forecasting information you were orignially seeking?]
    
    F --> J{No} 
    F --> I{Yes}
    
    J --> n17[Continue collecting data.]
    I --> n16[Congrats!]
    
    A --> C{No}
    C --> E[Is data available for the target, spactial unit, and time resolution?]
    E --> G{Yes} 
    E --> H{No}
    
    G --> K[Is historical data available for the target of interest?]
    H --> L[Does the data exist, but are not easily accesible?]
    
    K --> n1{Yes} 
    K --> n2{No}
    
    n1 --> n18[Collect and store the available data.]
    n2 --> n19[Go back to STEP 1. Reevaluate the project.]
    
    L --> n3{Yes}
    L --> n4{No}
    
    n3 --> n20[Collect any available data.]
    n4 --> n21[Go back to STEP 1. Reevaluate the project.]

    n18 --> n5
    n20 --> n5
    
    n5[Are these data reported or available in a consistent manner? e.g., same data is available every week on the same day] --> n6{Yes}
    n5 --> n7{No}
    
    n7 --> n12[Determine the reporting timeline and look for consistancies.]
    n12 --> n5
     
    n6 --> n15[Establish a reporting timeline consistent with the time resolution.]
    n15 --> n8
    
    n8[Setup consistent downloads of the entire dataset, not just the current week.] --> n9[Does your data have reporting delays or completeness issues?]
    n9 --> n10{Yes}
    n9 --> n11{No}
    
    n10 --> n13[Plan for Nowcasting.]
    n11 --> n14[Continue to data formatting steps for forecasting.]
")
```


```{r eval=FALSE, include=FALSE}
# Decision Tree Approach to Public Health Forecasting 

                                ┌───────────────────────────────────────┐
                                │  Did you choose state-level forecasts │
                                │        for a particular pathogen?     │
                                └────────────────────┬──────────────────┘
                                                     │
                                         ┌───────────────────────┐
                                         ▼                       ▼
                                 ┌───────────────┐       ┌───────────────┐
                                 │      Yes      │       │       No      │
                                 └───────┬───────┘       └───────┬───────┘
                                         │                       │
                                         ▼                       ▼
          ┌────────────────────────────────────┐            ┌────────────────────────────────────┐
          │  Review available forecasting hubs │            │ Is data available for the target,  │
          │   providing state-level forecast.  │     ┌──────│  spatial unit, and time resolution │
          └─────────────────┬──────────────────┘     │      │  you have chosen?                  │
                            │                        │      │  (e.g., weekly flu hosp in NC)     │
                            │                        │      └──────────────────┬─────────────────┘
                            ▼                        │                         │
              ┌───────────────────────────────┐      │               ┌───────────────────┐
              │  Did you find the state-level │      │               ▼                   ▼
              │  forecasting information you  │      │           ┌───────┐           ┌───────┐
              │  were originally seeking?     │      │           │  YES  │           │  NO   │
              └───────────────┬───────────────┘      │           └───────┘           └───────┘
                              │                      │               │                   │
                              │                      │               │                   │
                    ┌───────────────────┐            │               │                   │
                    ▼                   ▼            │               │                   │
     ┌───────────────────┐          ┌─────────┐      │               │                   │
     │  YES → Fantastic! │          │  NO →   │──────┘               │                   │
     └───────────────────┘          └────┬────┘                      │                   │
                                                                     │                   │
                                                                     ▼                   ▼
                                        ┌──────────────────────────────┐            ┌─────────────────────────────┐
                                        │ Is historical data available │            │ Does the data exist, but    │
                                        │  for the target of interest? │            │  are not easily accessible? │
                                        └──────────────┬───────────────┘            └─────────────┬───────────────┘
                                                       │                                          │
                          ┌─────────────────────────────────┐                      ┌─────────────────────────────┐
                          ▼                                 ▼                      ▼                             ▼
          ┌──────────────────────┐     ┌───────────────────────────┐      ┌────────────────────┐      ┌───────────────────────────┐
          │         YES →        │     │           NO →            │      │       YES →        │      │           NO →            │
          │  [Collect and store  │     │     Go back to Step 1.    │      │ [Collect any       │      │   Go back to Step 1.      │ 
          │  the available data] │     │  Re-evaluate the project. │      │  available data]   │      │  Re-evaluate the project. │
          └──────────┬───────────┘     └───────────────────────────┘      └──────────┬─────────┘      └───────────────────────────┘
                     │                                                               │
                     │                                                               │
                     └───────────────────────────────┬───────────────────────────────┘
                                                     │
                                                     │
                      ┌──────────────────────────────────────────────────────────────┐
                      │ Are these data reported or available in a consistent manner? │────────────────┐
                      │  (e.g., same data is available every week on the same day)   │                │
                      └──────────────────────────────┬───────────────────────────────┘                │
                                                     │                                                │
                         ┌────────────────────────────────────────────────────────┐                   │
                         │                                                        │                   │
                         │                                                        │                   │
                         ▼                                                        ▼                   │
           ┌──────────────────────────┐                              ┌──────────────────────────┐     │
           │            YES →         │                              │           NO →           │     │
           │ Establish a reporting    │                              │  Determine the reporting │─────┘
           │ timeline consistnet with │                              │  timeline and look for   │
           │ resolution.              │                              │  consistencies.          │
           └─────────────┬────────────┘                              └──────────────────────────┘
                         │                    
                         ▼                       
  ┌──────────────────────────────────────────────────────┐
  │   Setup consistent downloads of the entire dataset   │
  │ (not just the ci=urrent week - e.g., versioned data) │                
  └──────────────────────┬───────────────────────────────┘  
                         │                       
                         │
      ┌─────────────────────────────────────────┐                ┌────────────────────────┐
      │   Does your data have reporting delays  │────────────────│          YES →         │
      │       and completeness issues?          │                │  (Plan for Nowcasting) │
      └─────────────────────────────────────────┘                └────────────────────────┘

```

## National and State Level Forecasting Hubs:
FluSight [guide](https://happygitwithr.com/https-pat) 

MetroCast [website](information.https://docs.github.com/en/authentication/keeping-your-account-and-data-secure/managing-your-personal-access-tokens)

A full list of real-time collaborative public health hubs can be found [here](https://hubverse.io/community/hubs.html#real-time-collaborative-public-health-hubs).



## Next steps:

* For NHSN data, return to Get Started and use the get_data function to run forecasts 
* To forecast local surveillance data, follow [these](external_data.html) steps for formatting.


For more information, see the documentation and help pages.

